Analysis of the inference of ratings and rankings on Higher Order Networks with complex topologies

### 2025-09-05 draft/draft.tex
progress

Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$ to be rated and ranked.
Suppose that for each distinct pair of teams $i$ and $j$, a pairwise comparison score $\psi_{ij}=-\psi_{ji}$ is given.
Informally, $\psi_{ij}>0$ suggests that team $i$ should be rated higher than team $j$, while $\psi_{ij}<0$ suggests the opposite.

A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $w_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
$$
\sum_{i<j} \bigl(\psi_{ij}-(w_i-w_j)\bigr)^2.
$$
This yields a well-posed optimization problem that admits an elegant reformulation in the language of cochains over simplicial complexes.
The resulting framework for the inference of ratings and rankings is known as {\em HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.

@@

@@

>>>>

Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$ to be rated and ranked.
Suppose that for each distinct pair of teams $i$ and $j$, a pairwise comparison score $f_{ij}=-f_{ji}$ is given.
Informally, $f_{ij}>0$ suggests that team $i$ should be rated higher than team $j$, while $f_{ij}<0$ suggests the opposite.

A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $w_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
$$
\sum_{i<j} \bigl(f_{ij}-(w_i-w_j)\bigr)^2.
$$
This yields a well-posed optimization problem that admits an elegant reformulation in the language of cochains over simplicial complexes.
The resulting framework for the inference of ratings and rankings is known as {\em HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.

@@

\subsection{Discrete calculus on simplicial complexes}

@@

Next, the simplicial complex is equipped with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.

The vector space of chains $C$ is equipped with the trivial inner product
$$
e_r.e_s = \delta_{rs}
$$
for each pair $r,s \in K$.
Using Einstein's implicit summation convention, an arbitrary chain $c$ in $C$ can be linearly decomposed in the canonical basis as $c=c^se_s$ where the $c^s$ are coefficients in $\RR$.
The vector space of chains admits the following decomposition in direct sums
$$
C = C_0\oplus C_1\oplus C_2
$$
where $C_k$ is the subspace of $C$ spanned by the basis of $k$-chains, i.e. by $\{e_s:s\in K_k\}$.

The covectors $f$ in the dual space $C^*$ of $C$ are called cochains.
The dual basis is denoted by $\{e^s:s\in K\}$ and satisfies
$$
e^r(e_s) = \delta^r{}_s
$$
for each $s,r\in K$.
An arbitrary cochain $f$ in $C^*$ can be linearly decomposed in the dual basis as $f=f_se^s$ where the $f_s$ are coefficients in $\RR$.
The vector space of cochains admits the following decomposition in direct sums
$$
C^* = C_0^*\oplus C_1^*\oplus C_2^*
$$
where $C_k^*$ is the subspace of $C^*$ spanned by the dual basis of $k$-cochains, i.e. by $\{e^s:s\in K_k\}$.

Riesz's theorem establishes that for any cochain $f\in C^*$, there is an unique chain in $C$ denoted by $f^{\flat}$ such that $f(q) = f^{\flat}.c$ for any chain $c\in C$.
This results, induces an inner product on on $C^*$ given by
$$
f.g = f^{\flat}.g^{\flat}
$$
for any pair of cochains $f,g\in C^*$.
The map $C^*\ni f\to f^{\flat}\in C$ is an isomorphism, whose inverse is denoted by $C\ni c\to c^{\sharp}\in C^*$.

A notion of orientation is imposed to the vector space of chains, by requering that
$$
e_s
=
e_{i_0...i_k}
=
(-1)^p
e_{j_0...j_k}
$$
for each simplex $s=\{i_0,...,i_k\}$ of $K$
when $0\leq i_0<i_1<...<i_k\leq n$ and $p$ is the parity of the permutation taking the sequence $i_0...i_k$ into the sequence $j_0...j_k$.

Thanks to the notion of orientation, a boundary linear operator $\partial:C\to C$ can be defined by its action on the vectors of the canonical basis, i.e. by
$$
\partial e_{i_0...i_k}
=
\sum_{j=0}^k
(-1)^j
e_{i_0...i_{j-1}i_{j+1}...i_k}
$$
for any $e_{i_0...i_k}\in C$.
Note that $\partial(C_k) \subseteq C_{k-1}$, where $C_{-1}=\{0\}$ denotes the trivial subspace of $C$ of dimension 0.

Since the application of any cochain $f$ to any chain $c$ is a bilinear form, the coboundary linear operator $d:C^*\to C^*$ is univocally defined by the relation
$$
df(c) = f(\partial c)
$$
for any $f\in C^*$ and $c\in C$.
Note, $d(C_k^*)\subset C_{k+1}^*$ where $C_3^*=\{0\}$ denotes the trivial subspace of $C^*$ of dimension 0.

Since the inner products on $C$ and $C^*$ are also bilinear forms, then the dual boundary $\partial^*:C\to C$ and the dual coboundary $d^*:C^*\to C^*$ linear operators can be univocally defined by
$$
\partial^*c.q
=
c.\partial q
$$
and
$$
d^*f.g
=
f.d g,
$$
for any chains $c,q\in C$ and any cochains $f,g\in C^*$, respectively.
Note, $\partial^*(C_{k-1})\subseteq C_k$ and $d^*(C_{k+1}^*)\subseteq C_k^*$.
